Individual and Differential Harm in Redistricting

### 2022-05-23 paper/paper.Rmd
rough draft intro

---
title: Individual and Differential Harm in Redistricting
authors:
  - name: Cory McCartan
    thanks: |
      To whom correspondence should be addressed.
      The authors thank participants of the Harvard American Politics
      Research Workshop and Imai Research Group  for helpful comments and discussion.
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  We propose a unified framework for evaluating the impact of a districting plan on voters overall and in different groups. Considering the harm done to individual voters by their inability to elect their chosen candiate, and taking into account the space of counterfactual possibilities, we derive a set of measures that flexibly adapts to local geographic constraints, different types of elections, and different groups of study. We provide recommendations for estimating these measures from historical election returns and discuss connections to commonly used redistricting metrics. Finally, we apply our framework to study partisan gerrymandering in New Jersey, voting rights litigation in Alabama, and the racial impacts of city council district boundaries in Boston.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
biblio-style: chicago
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
editor_options:
  markdown:
    wrap: sentence
---

@@

-   why counterfactuals?

    -   many factors influence the performance and qualities of an electoral system

    -   understanding how different components of the system contribute requires reaching in and manipulating those components

    -   just like causal inference, you are inherently interested in the counterfactual where some part of the electoral system is different

-   why representational harm?

    -   lit cites on why districts are good

    -   district-based electoral system working well if people are represented by the candidate of their choice

    -   this motivates *representational harm*

    -   we argue that harm is actually a great unifying framework for studying district-based electoral systems

-   what makes a good redistricting plan?

-   on the partisan side, lots of attention on the fairness to parties

    -   partisan symmetry [@katz2020]

    -   seats-votes curve

-   on the racial side, ... ?
    (VRA focus? looking at race of candidates vs choice? what? I don't know)

-   courts, on the other hand, have focused on individual standing [@1964reynolds; @1962baker]

-   **we argue that individual standing/harm is an excellent lens to unify the study and measurement of districting plans**

-   core idea: counterfactuals + voter choice.
    Is a voter more likely to be represented by the candidate of their choice under a different districting plan?
    If so, they've been harmed.

    -   modern simulation methods allow us to produce realistic counterfactuals

    -   by analyzing how harms fall differentially along party, race, ideological, etc. lines, we can connect individual harm to more global measures based on parties, etc.

<!--# Yet, this suffers from a core inflexibility: it does not allow for variation in districting. Districting is the fundamental question of district-based democracies. For whom can you vote? Who do you vote with? Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation. The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual. -->

@@

Section \ref{sec:motivation} motivates our framework for measuring harm and illustrates the key concepts in the context of partisan gerrymandering for three simple examples.
Section \ref{sec:indiv-harm} formalizes the intuition of Section \ref{sec:motivation} and defines a general measure of individual and differential harm.
Section \ref{sec:estimating} discusses estimating these measures from observed election results, and provides what we believe is a sensible and efficient way to calculate the measures in the special case of studying partisan gerrymandering.
Section \ref{sec:connections} connects our proposed measure to existing practices for analyzing racial and partisan gerrymandering.
Sections \ref{sec:nj}, \ref{sec:alabama}, and \ref{sec:boston} apply the proposed measures to partisan gerrymandering of New Jersey congressional districts, voting rights litigation for Alabama congressional districts, and racial impacts of district boundaries on Boston City Council primary elections.
Section \ref{sec:discussion} concludes.

@@

There is also disagreement between the measures about the natural bias of districting plans in New Jersey: the median sampled plan has a partisan bias of $0.047$, indicating a natural Republican bias, but the median mean-median score is just $0.0069$, and the median differential partisan harm is $-0.13$, indicating a natural Democratic bias.
Which is correct?
Democrats win 8.47 seats on average across the simulations, which is more than the 6.79 seats that would be expected under a completely proportional outcome.

@@

To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^10] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.

>>>>

---
title: Individual and Differential Harm in Redistricting
authors:
  - name: Cory McCartan
    thanks: |
      To whom correspondence should be addressed.
      The authors thank participants of the Harvard American Politics
      Research Workshop and Imai Research Group  for helpful comments and discussion.
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  We propose a unified framework for evaluating the impact of a districting plan on voters overall and in different groups. Considering the harm done to individual voters by their inability to elect their chosen candiate, and taking into account the range of counterfactual possibilities, we derive a set of measures that flexibly adapts to local geographic constraints, different types of elections, and different groups of study. We provide recommendations for estimating these measures from historical election returns and discuss connections to commonly used redistricting metrics. Finally, we apply our framework to study partisan gerrymandering in New Jersey, voting rights litigation in Alabama, and the racial impacts of city council district boundaries in Boston.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
biblio-style: chicago
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
editor_options:
  markdown:
    wrap: sentence
---

@@

For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
Partisan gerrymandering, whether in the form of rotten boroughs or the salamander-shaped districts of Elbridge Gerry [cite], is perhaps the most consistently visible manifestation of this tendency.
The academic study of partisan gerrymandering has largely centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@katz2020, need more].
In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [cite].
The Voting Rights Act of 1965 (VRA) provided a framework for legal challenges to racial gerrymanders that has also shaped the academic literature, which is characterized by a focus on identifying the voting preferences of racial groups and observing the extent to which these are realized or frustrated through district boundaries and other features of the electoral system [cites].
At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the electoral system may be more subtle or may favor a narrower or less well-defined group [cite].
Local redistricting has been less extensively studied, in part for this reason and in part because of data limitations [but see \_\_\_\_].

No matter the source or context, all of these cases involve the concentration or dilution of voting power to favor or disfavor a particular group.
This paper proposes a new framework of *individual and differential harm* for the unified study of all forms of gerrymandering by way of measuring this concentration and dilution.

The harm framework combines two key ideas.
First, we operate at the level of the individual voter, and examine whether each voter is represented by the candidate of their choice.
Doing so both unifies our approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public [cites on why districts are good].

Second, the proposed framework moves beyond a single electoral system under study to also consider *counterfactual* electoral systems.
Many factors influence the performance and qualities of an electoral system.
Without a well-defined comparison system, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain.
As we demonstrate, modern redistricting simulation tools can provide a realistic set of counterfactuals that make these kinds of comparisons possible.

The combination of individual voter choices and counterfactual electoral systems defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different electoral system.
And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.

This framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker].
This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidates [case cite].

@@

In Section \ref{sec:motivation}, we motivate our framework for measuring harm and illustrate the key concepts with three simple examples.
Section \ref{sec:indiv-harm} formalizes the intuition of Section \ref{sec:motivation} and defines a general measure of individual and differential harm.
Section \ref{sec:estimating} discusses estimating these measures from observed election results, and provides what we believe is a sensible and efficient way to calculate the measures in the special case of studying partisan gerrymandering.
Section \ref{sec:connections} connects our proposed measure to existing practices for analyzing racial and partisan gerrymandering.
Sections \ref{sec:nj}, \ref{sec:alabama}, and \ref{sec:boston} apply the proposed measures to partisan gerrymandering of New Jersey congressional districts, voting rights litigation for Alabama congressional districts, and racial impacts of district boundaries on Boston City Council primary elections.
Section \ref{sec:discussion} concludes.

@@

There is also disagreement between the measures about the natural bias of districting plans in New Jersey: the median sampled plan has a partisan bias of $0.047$, indicating a natural Republican bias, but the median mean-median score is just $0.0069$, and the median differential partisan harm is $-0.13$, indicating a natural Democratic bias.
Which is correct?
It depends on what one means by "natural bias," though we submit that differential harm is one good way to measure it.
Democrats do win 8.47 seats on average across the simulations, which is more than the 6.79 seats that would be expected under a completely proportional outcome.

@@

To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^10] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail.
